HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-04 scnn.tex
add some comments

Convolutional Neural Networks (CNNs) are now a cornerstone of the Deep Learning toolbox and have led to many breakthroughs in Artificial Intelligence. Despite their successes, they still are mostly developed and applied to regular Euclidean domains such as those supporting images, audio, or video. For many applications however, the data domain is not regular.
As a response to that fact, we are witnessing an increasing interest in the generalization of CNNs to graphs. Graphs are versatile data structures which can represent pairwise relationships between objects or act as a discrete representation of a continuous manifold.
In this paper we take the latter view and represent a sampled sphere as a graph, for which we define the convolution and pooling operations.
Spherical maps appear in many applications, such as the study of phenomena on the Earth for meteorologists or on the sky for cosmologists.
A popular format for spherical maps is Healpix, which uses an equal-area, isolatitude sampling. Hence, we present a spherical CNN for analysis of full and partial Healpix maps, which we call HealpixNet.
While convolution on arbitrary graphs has mostly been sorted out, a general pooling operation has not been discovered so far. As such, while HealpixNet is a special case of a graph CNN, it is tailored to the Healpix sampling of the sphere.
This approach is computationally much more efficient than using spherical harmonics to perform convolutions, an alternative that as been explored previously.
% The filters learned by the HealpixNet are restricted to being radial.
% Michaël: not necessary in the abstract?
We demonstrate the method on a classification problem of mass maps from two cosmological models. A standard benchmark for that problem is to train an SVM on the histograms of map pixel values. Another is to use the power spectral density instead.
HealpixNet achieves \todo{xx\%} better accuracy than those baselines, is more robust to noise, and is \todo{faster?}.
Code and examples are available at \todo{\url{https://github.com/xxx}}.
\end{abstract}

@@

\todo{@Tomek: can you explain why we care about parts of the sphere? (that is an advantage of the graph model over the spherical harmonics)}
\todo{@Tomek: How is the data acquired? Do telescopes and satellites take measurements according to the Healpix sampling? (it is also an advantage of our method to be agnostic to the sampling, which might be useful to researchers working with missing data or different models)}

So far these algorithms have mostly been demonstrated on Euclidean domains, such as flat images.
As such, perhaps the most straightforward approach to the analysis of spherical data is to divide it into small chunks and project those on flat 2D surfaces.
While this mostly works, as demonstrated \todo{in ??? for example}, it is more natural and accurate to work directly on the sphere.

@@

To overcome those shortcomings, we introduce an efficient spherical CNN formulation which uses (i) convolutions on graphs and (ii) hierarchical pooling. The main idea is to represent the sphere $S^2$, a 2-dimensional manifold embedded in $\mathbb{R}^3$, as a graph of connected pixels.
Our approach uses the graph CNN formulation introduced by \citet{defferrard2016convolutional}, which is efficient and can adapt to any sampling of the sphere. The convolution operation costs $\mathcal{O}(n)$ operations, where $n$ is the number of pixels. That is the lowest possible complexity for convolution without approximations (e.g. by sketching).
The flexibility of modeling the data domain with a graph allows, for example, to easily model data which only spans a part of the sphere, or data which is not uniformly sampled.
Our hierarchical pooling strategy exploits a hierarchical pixelization of the manifold to analyze the data at multiple scales. Such hierarchical pixelizations of the sphere have been well studied.

@@

We give a practical demonstration of the application of our package to cosmological model discrimination using convergence maps, similar to \citep{schmelze2017cosmologicalmodel}.
In a simplified scenario, we classify convergence maps on parts of a sphere into two cosmological models.
These models were designed to have the same power spectrum in range $\ell < 1000$.
We compare the performance of the spherical CNN to a baseline algorithm, an SVM classifier which takes pixel histograms or Power Spectrum Densities (PSD) of these maps as input.
The comparison is made as a function of noise level on convergence maps and area of the sphere used. Results show that our model is better at discriminating the maps, is more resilient to noise, and \todo{is faster?}.

@@

% About the the different sphere sampling.
% \begin{itemize}
% 	\item Define the spherical grid?
% 	\item reference the "cubed sphere"~\cite{ronchi1996cubed}?
% \end{itemize}

@@

\subsection{Spherical convolutions}
\label{sec:spherical_conv}
The major challenge to define Convolution Neural Network (CNN) onto a
spherical domain is to handle a convolution that is suitable for this irregular sampling domain.
Three approaches exist to address this issue.

@@

A second direction has been followed in~\citet{boomsma2017spherical}. The idea is
to use the traditional 2-dimensional convolution on an irregular grid defined on the
sphere. In the paper, two grids are used: the geodesic grid \todo{isn't it the equirectangular grid?} and the ``cubed
sphere'' grid defined by~\citet{ronchi1996cubed}. Because this
approach is based on the traditional 2-dimensional convolution, it is very likely to be the
most computationally efficient one. However, it suffers that it can only be used using very
specific grid sampling sets which are NOT including HealPix. Furthermore, the
grid requirement makes it impossible for the convolution to capture the
spherical properties of the domain, i.e. the "cubed sphere” sampling is by
definition adapted to a cube and not to a sphere.
\todo{border effects?}
% \nati{Michael: Under some very
% specific hypothesis, this second approach is a particular case of our method,
% i.e. 1) our method with a stupid sampling, 2) assumption that graph convolution
% on a grid == 2d convolution. Do you think we should mention that? I think we
% should not.}
% I think we should!

@@

\todo{Cool to have a global illustration of the network (CNN like)}

\subsection{Convolutions on graphs}
\assign{Michaël}

@@

\todo{Maybe we just describe convolutions on sphere and graph, i.e: do we really need this two subsections? Maybe simply merge them?}
\subsection{Convolutions on manifolds}

>>>>

Convolutional Neural Networks (CNNs) are now a cornerstone of the Deep Learning toolbox and have led to many breakthroughs in Artificial Intelligence. Despite their successes, they still are mostly developed and applied to regular Euclidean domains such as those supporting images, audio, or video. For many applications however, the data domain is not regular.
As a response to that fact, we are witnessing an increasing interest in the generalization of CNNs to graphs. Graphs are versatile data structures which can represent pairwise relationships between objects or act as a discrete representation of a continuous manifold.
In this paper we take the latter view and represent a sampled sphere as a graph, for which we define the convolution and pooling operations.
Spherical maps appear in many applications, such as the study of phenomena on the Earth for meteorologists or on the sky for cosmologists.
A popular format for spherical maps is Healpix, which uses an equal-area, isolatitude sampling. Hence, we present a spherical CNN for analysis of full and partial Healpix maps, which we call HealpixNet.
\nati{I am not that we should talk about pooling in the abstract. Another angle would be to say that HealPix sampling is well suited for spherical CNN with graphs as a) it is very regular => the graph capture well the spherical Manifold structure and b) pooling is natural.} 
While convolution on arbitrary graphs has mostly been sorted out, a general pooling operation has not been discovered so far. 
As such, while HealpixNet is a special case of a graph CNN, it is tailored to the Healpix sampling of the sphere.
This approach is computationally much more efficient than using spherical harmonics to perform convolutions, an alternative that as been explored previously.
% The filters learned by the HealpixNet are restricted to being radial.
% Michaël: not necessary in the abstract?
We demonstrate the method on a classification problem of mass maps from two cosmological models. 
\nati{Should we specify what are the baselines? I think not.}
A standard benchmark for that problem is to train an SVM on the histograms of map pixel values. Another is to use the power spectral density instead.
HealpixNet achieves \todo{xx\%} better accuracy than those baselines, is more robust to noise, and is \todo{faster?}.
Code and examples are available at \todo{\url{https://github.com/xxx}}.
\end{abstract}

@@

\todo{@Tomek: can you explain why we care about parts of the sphere? (that is an advantage of the graph model over the spherical harmonics)}
\todo{@Tomek: How is the data acquired? Do telescopes and satellites take measurements according to the Healpix sampling? (it is also an advantage of our method to be agnostic to the sampling, which might be useful to researchers working with missing data or different models)}
\nati{@Tomek: I think what Michael has in mind is that the graph approach is very versatile and extendable. It can be used with other samplings (even irregular ones) or only part of the sphere. This flexibility might be an advantage to mention. In particular, compared to~\citet{cohen2018spherical} it allows us to do only do the required computation when only a part of the sphere is considered.}

So far these algorithms have mostly been demonstrated on Euclidean domains, such as flat images.
As such, perhaps the most straightforward approach to the analysis of spherical data is to divide it into small chunks and project those on flat 2D surfaces.
While this mostly works, as demonstrated \todo{in ??? for example}, \nati{@Tomek, do you have a citation for that? Otherwise, we need to reformulate this.} it is more natural and accurate to work directly on the sphere.

@@

To overcome those shortcomings, we introduce an efficient spherical CNN formulation which uses (i) convolutions on graphs and (ii) hierarchical pooling. The main idea is to represent the sphere $S^2$, a 2-dimensional manifold embedded in $\mathbb{R}^3$, as a graph of connected pixels.
Our approach uses the graph CNN formulation introduced by \citet{defferrard2016convolutional}, which is efficient and can adapt to any sampling of the sphere. Using a k-nearest neigboors graph, the convolution operation costs $\mathcal{O}(n)$ operations, where $n$ is the number of pixels. That is the lowest possible complexity for convolution without approximations (e.g. by sketching).
The flexibility of modeling the data domain with a graph allows, for example, to easily model data which only spans a part of the sphere, or data which is not uniformly sampled.
Our hierarchical pooling strategy exploits a hierarchical pixelization of the manifold to analyze the data at multiple scales. Such hierarchical pixelizations of the sphere have been well studied.

@@

We give a practical demonstration of the application of our package to cosmological model discrimination using convergence maps, similar to \citep{schmelze2017cosmologicalmodel}.
In a simplified scenario, we classify convergence maps on parts of a sphere into two cosmological models.
These models were designed to have the same power spectrum in range $\ell < 1000$.
We compare the performance of the spherical CNN to a baseline algorithm, an SVM classifier which takes pixel histograms or Power Spectrum Densities (PSD) of these maps as input.
The comparison is made as a function of noise level on convergence maps and area of the sphere used. Results show that our model is better at discriminating the maps, is more resilient to noise, and \todo{is faster?}.
\nati{I do not think we should argue for speed gain in this paper. We are probably much more efficient than scnns with the convolution done in the spectral domain. However compared to tradiationl methods, we are in most of the case slower.}

@@

@@

\subsection{Spherical convolutions and spherical CNNs}
\label{sec:spherical_conv}
The major challenge to define Convolution Neural Network (CNN) onto a
spherical domain is to handle a convolution that is suitable for this irregular sampling domain.
Three approaches exist to address this issue.

@@

A second direction has been followed in~\citet{boomsma2017spherical}. The idea is
to use the traditional 2-dimensional convolution on an irregular grid defined on the
sphere. In the paper, two grids are used: the equirectangular grid and the ``cubed
sphere'' grid defined by~\citet{ronchi1996cubed}. \nati{Do we need a plot of the equirectangular and cubed sphere sampling?} Because this
approach is based on the traditional 2-dimensional convolution, it is very likely to be the
most computationally efficient one. However, it suffers that it can only be used using very
specific grid sampling sets which are NOT including HealPix. Furthermore, the
grid requirement makes it impossible for the convolution to capture the
spherical properties of the domain, i.e. the "cubed sphere” sampling is by
definition adapted to a cube and not to a sphere.
\todo{border effects?}
% \nati{Michael: Under some very
% specific hypothesis, this second approach is a particular case of our method,
% i.e. 1) our method with a stupid sampling, 2) assumption that graph convolution
% on a grid == 2d convolution. Do you think we should mention that? I think we
% should not.}
% I think we should!

@@

\subsection{Convolutions on graphs}
\todo{Maybe we just describe convolutions on sphere and graph, i.e: do we really need all these three subsections? Maybe simply merge them?}

\nati{I think that citing Bruna is to much... According to me, the goal of this subsection should not be to cite everything that exist. But should focus on
1) be fair and say that they are other graph convolutions that can be used, 2) show some connections between graph CNNs ang group equivariant CNNs (Done), and 3) suggest that they are possible extensions of what we do.}

\assign{Michaël}

@@

\subsection{Convolutions on manifolds}
